---
title: "TABio_paper_analysis"
output: html_document
date: "2024-08-15"
---


# Two-arm Biofeedback Study

## Background

This study compares a control group with a training group on two virtual reality
games aiming at teaching participants a breathing technique and subsequently to apply
this technique to a stressful environment 


### Study Design

Both groups came in over two sessions. In the first session both groups took a baseline 
physiological measurement (seated at rest) and then went on to a virtual reality stressor (horror) experience. Then the training groups did two sessions of the slow breathing training.

In the second session, training group participants did two more sessions of the slow breathing 
training. Both groups then took part in the biofeedback stressor, which asked them to control their
breathing to keep their heart rate low to avoid detection by a monster. 

The physiological measurements reported here are Heart Rate (HR), heart
rate variability as indexed by SDNN, and respiration rate (resp).

Initial analysis was conducted using a paired t-test for the HR and SDNN
difference from baseline to the biofeedback stressor.



``` {r results='hide', message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(nlme)
library(ggplot2)
library(ggpubr)
library(rstatix)
library(reshape2)
library(dplyr)
library(psych)



```

``` {r echo = FALSE}

training_t2 <- read.csv("/Users/luciedaniel-watanabe/Desktop/NT:PhD/TABio/TABio Analysis/data/training_t2_CLEAN.csv")
control_t2 <- read.csv("/Users/luciedaniel-watanabe/Desktop/NT:PhD/TABio/TABio Analysis/data/control_t2_CLEAN.csv")
control_t1<- read.csv("/Users/luciedaniel-watanabe/Desktop/NT:PhD/TABio/TABio Analysis/data/control_t1_CLEAN.csv")
training_t1 <- read.csv("/Users/luciedaniel-watanabe/Desktop/NT:PhD/TABio/TABio Analysis/data/training_t1_CLEAN.csv")


#cbPalette <- c( "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
cbPalette <- c( "#D55E0095", "#0072B295")

##checking the ids match for both groups 
#training_t1$ids == training_t2$ids
#ßcontrol_t1$ids == control_t2$ids

#rm <- sample(training_t1$ids, 1) 
### initially used this to randomly choose a participant from 
## the training group to exclude to make the groups even sizes - 
## the initial exclusion was 9361 so it has been hardcoded 
###in for reproducibility 

training_t1[which(training_t1$ids == 9361),] <- NA
training_t2[which(training_t2$ids == 9361),] <- NA

training_t2 <- training_t2[complete.cases(training_t2),]
training_t1 <- training_t1[complete.cases(training_t1),]



training_t2$hr_diff <- training_t2$dungeon_HR - training_t1$baseline_HR
control_t2$hr_diff <- control_t2$dungeon_HR - control_t1$baseline_HR

training_t2$sdnn_diff <- training_t2$dungeon_sdnn - training_t1$baseline_sdnn
control_t2$sdnn_diff <- control_t2$dungeon_sdnn - control_t1$baseline_sdnn

options(scipen=999)

################################################
## T-tests 
################################################


t.test(training_t2$hr_diff, control_t2$hr_diff)
t.test(training_t2$sdnn_diff, control_t2$sdnn_diff)


```

### Two-way repeated measures ANOVA for HR 

We then did a two-way repeated measures ANOVA with the horror virtual reality from
session 1 and the biofeedback stressor from session 2 to examine the differences 
between control and training groups 

``` {r, results='hide', message=FALSE, warning=FALSE, echo=FALSE, fig.show='hide'}


################################################
### BL normalised HR repeated measures ANOVA ###
################################################



training_melt <- data.frame(ids = rep(training_t1$ids,2),
                            time = c(rep("intruder", 27), rep("stressor",27)),
                            HR = c(training_t1$horror_HR - training_t1$baseline_HR,
                                   training_t2$hr_diff),
                            group = rep("training", 54))

control_melt <- data.frame(ids = rep(control_t1$ids,2),
                           time = c(rep("intruder", 27), rep("stressor",27)),
                           HR = c(control_t1$horror_HR - control_t1$baseline_HR,
                                  control_t2$hr_diff),
                           group = rep("control", 54))


melt.aov <- rbind(control_melt, training_melt)

#rm(training_melt, control_melt)

melt.aov$time <- factor(melt.aov$time, levels=c("intruder", "stressor"))
melt.aov$group <- factor(melt.aov$group)


#test normality 
melt.aov %>%
  group_by(time, group) %>%
  shapiro_test(HR)

ggqqplot(melt.aov, "HR", ggtheme = theme_bw()) +
  facet_grid(time ~ group)

#test homogeneity of variance
melt.aov %>%
  group_by(time) %>%
  levene_test(HR ~ group)

#test homogeneity of covariance
box_m(melt.aov[, "HR", drop = FALSE], melt.aov$group)


```

``` {r, echo = FALSE,  warning=FALSE}


# Two-way mixed ANOVA test
res.aov <- anova_test(
  data = melt.aov, dv = HR, wid = ids,
  between = group, within = time
)
get_anova_table(res.aov)

# Effect of time at each level of group
one.way <- melt.aov %>%
  group_by(group) %>%
  anova_test(dv = HR, wid = ids, within = time) %>%
  get_anova_table() %>%
  adjust_pvalue(method = "fdr")
one.way

##calculate the means for control and training

mean(control_t1$horror_HR)
mean(control_t2$dungeon_HR)


melt.aov %>% 
  group_by(group, time) %>% 
  summarise(hr_mean = mean(HR)) -> melt.aov2


melt.aov %>% 
  group_by(group, time) %>% 
  summarise(sd = sd(HR)) -> melt.aov3

melt.aov2$sd = melt.aov3$sd
melt.aov2$sd = melt.aov2$sd / sqrt(27)

rm(melt.aov3)


hr_aov_plot<- ggplot(data = melt.aov2) +
  aes(x = time, y = hr_mean, color = group) +
  geom_line(aes(group = group), linewidth = 1.2) +
  scale_color_manual(values=cbPalette)+
  labs(title = "",
       x = "Condition",
       y = "HR ",
       fill = "group")+
  geom_point()+
  ggtitle("HR in Pre-Training Stressor (Intruder)\n and Post-Training Stressor (Stressor)")+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.title = element_text(margin=margin(0,0,13,0))) +
  theme(plot.title=element_text(face="bold")) +
  theme(plot.title=element_text(size=12)) +
  theme(axis.text = element_text(colour = "#00000095")) +
  theme(axis.title.x = element_text(size = 13)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(axis.text=element_text(size=12))+
  theme(axis.title.x = element_text(margin=margin(13,0,0,0))) +
  theme(axis.title.y = element_text(margin=margin(0,13,0,0))) +
  theme(legend.text = element_text(size=18))+
  theme(legend.key.size = unit(2, 'cm'))+
  geom_errorbar(aes(ymin = hr_mean-sd, ymax= hr_mean+sd),width=.2,
                position=position_dodge(0.05)) +
  scale_y_continuous(n.breaks=5) 

hr_aov_plot  + guides(colour=FALSE)


rm(melt.aov, melt.aov2, model.aov, res.aov, one.way)

```

### Two-way repeated measures ANOVA for SDNN 


```{r, results='hide', message=FALSE, warning=FALSE, echo=FALSE, fig.show='hide'}
###################################################
### BL normalised SDNN repeated measures ANOVA ###
##################################################

training_sdnn_melt <- data.frame(ids = rep(training_t1$ids,2),
                                 time = c(rep("intruder", 27), rep("stressor",27)),
                                 SDNN = c(training_t1$horror_sdnn - training_t1$baseline_sdnn,
                                          training_t2$sdnn_diff),
                                 group = rep("training", 54))

control_sdnn_melt <- data.frame(ids = rep(control_t1$ids,2),
                                time = c(rep("intruder", 27), rep("stressor",27)),
                                SDNN = c(control_t1$horror_sdnn - control_t1$baseline_sdnn,
                                         control_t2$sdnn_diff),
                                group = rep("control", 54))

melt.aov <- rbind(control_sdnn_melt, training_sdnn_melt)

rm(control_sdnn_melt, training_sdnn_melt)


melt.aov$time <- factor(melt.aov$time, levels=c("intruder", "stressor"))
melt.aov$group <- factor(melt.aov$group)


#test normality dATA IS NOT NORMAL
melt.aov %>%
  group_by(time, group) %>%
  shapiro_test(SDNN)

ggqqplot(melt.aov, "SDNN", ggtheme = theme_bw()) +
  facet_grid(time ~ group)

#test homogeneity of variance
melt.aov %>%
  group_by(time) %>%
  levene_test(SDNN ~ group)

#test homogeneity of covariance
box_m(melt.aov[, "SDNN", drop = FALSE], melt.aov$group)

# Two-way mixed ANOVA test
res.aov <- anova_test(
  data = melt.aov, dv = SDNN, wid = ids,
  between = group, within = time
)
get_anova_table(res.aov)


# Pairwise comparisons between group levels
pwc <- melt.aov %>%
  group_by(time) %>%
  pairwise_t_test(SDNN ~ group, p.adjust.method = "bonferroni")
pwc

melt.aov <- melt.aov %>% 
  group_by(group) %>% 
  mutate(mean = mean(SDNN))

melt.aov <- melt.aov %>% 
  group_by(time) %>% 
  mutate(time_mean = mean(SDNN))

# Effect of time at each level of exercises group
one.way <- melt.aov %>%
  group_by(group) %>%
  anova_test(dv = SDNN, wid = ids, within = time) %>%
  get_anova_table() %>%
  adjust_pvalue(method = "fdr")
one.way


# Pairwise comparisons between time points at each group levels
# Paired t-test is used because we have repeated measures by time
pwc2 <- melt.aov %>%
  group_by(group) %>%
  pairwise_t_test(
    SDNN ~ time, paired = TRUE, 
    p.adjust.method = "bonferroni"
  ) %>%
  select(, -p) # Remove details

print(pwc2)



bxp <- ggboxplot(
  melt.aov, x = "time", y = "SDNN",
  color = "group", palette= cbPalette,
  ylab = "SDNN", xlab = "Condition") +   
  theme_minimal()+
  theme(axis.text = element_text(colour = "#00000095")) +
  theme(axis.title.x = element_text(face="bold")) +
  theme(axis.title.y = element_text(face="bold")) +
  theme(axis.title.x = element_text(size = 12)) +
  theme(axis.title.y = element_text(size = 12)) +
  theme(axis.text=element_text(size=11))+
  theme(axis.title.x = element_text(margin=margin(13,0,0,0))) +
  theme(axis.title.y = element_text(margin=margin(0,13,0,0))) +
  theme(legend.text = element_text(size=18))+
  theme(legend.key.size = unit(2, 'cm'))+
  scale_y_continuous(n.breaks=5) 

bxp



melt.aov %>% 
  group_by(group, time) %>% 
  summarise(sdnn_mean = mean(SDNN)) -> melt.aov2


melt.aov %>% 
  group_by(group, time) %>% 
  summarise(sd = sd(SDNN)) -> melt.aov3

melt.aov2$sd = melt.aov3$sd
melt.aov2$sd = melt.aov2$sd / sqrt(27)

rm(melt.aov3)



sdnn_aov_plot <- ggplot(data = melt.aov2) +
  aes(x = time, y = sdnn_mean, color = group) +
  geom_line(aes(group = group), linewidth = 1.2) +
  scale_color_manual(values=cbPalette)+
  labs(title = "",
       x = "Condition",
       y = "cvSDNN ",
       fill = "group")+
  geom_point() +
  ggtitle("cvSDNN in Pre-Training Stressor (Intruder)\n and Post-Training Stressor (Stressor)")+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.title = element_text(margin=margin(0,0,13,0))) +
  theme(plot.title=element_text(face="bold")) +
  theme(plot.title=element_text(size=12)) +
  theme(axis.text = element_text(colour = "#00000095")) +
  theme(axis.title.x = element_text(size = 13)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(axis.text=element_text(size=12))+
  theme(axis.title.x = element_text(margin=margin(13,0,0,0))) +
  theme(axis.title.y = element_text(margin=margin(0,13,0,0))) +
  theme(legend.text = element_text(size=18))+
  theme(legend.key.size = unit(2, 'cm'))+
  geom_errorbar(aes(ymin = sdnn_mean-sd, ymax= sdnn_mean+sd),width=.2,
                position=position_dodge(0.05)) +
  scale_y_continuous(n.breaks=5) 




rm(melt.aov, melt.aov2, bxp, pwc, pwc2, one.way, res.aov)

```
``` {r echo = FALSE}
sdnn_aov_plot + theme(legend.title = element_blank())

```

## Respiration analysis

We also examined the difference in respiration between the control and
training groups with a T test 


``` {r, echo = FALSE, warning = FALSE, message = FALSE}

###################################################
############## Respiration analysis ############## 
##################################################


controlResp <- read.csv("/Users/luciedaniel-watanabe/Desktop/NT:PhD/TABio/TABio Analysis/data/controlResp.csv")
trainResp <- read.csv("/Users/luciedaniel-watanabe/Desktop/NT:PhD/TABio/TABio Analysis/data/trainingResp.csv")

trainResp$diff <- trainResp$dungeonBreathTrain - trainResp$baselineBreathTrain
controlResp$contDiff <- controlResp$dungeonBreathCont - controlResp$baselineBreathCont


# rm <- sample(trainResp$ids, 4) 
# Initially I selected 4 random datasets from the training group to exclude
#these were 6782, 9361, 6924 and 4486
# for replicability I'm now manually removing those participants
# so we don't randomly remove different participants each time and alter results

trainResp[which(trainResp$ids == 6782),] <- NA
trainResp[which(trainResp$ids == 9361),] <- NA
trainResp[which(trainResp$ids == 6924),] <- NA
trainResp[which(trainResp$ids == 4486),] <- NA

trainResp <- trainResp[complete.cases(trainResp),]


mean(trainResp$diff)
mean(controlResp$contDiff)


t.test(controlResp$contDiff, trainResp$diff)


#### plotting difference in resp change
### between control and training groups 


colnames(trainResp)[2] <- "baseline" 
colnames(trainResp)[3] <- "stressor" 
colnames(controlResp)[2] <- "baseline" 
colnames(controlResp)[3] <- "stressor" 



resp_plot <- data.frame(controlBL = controlResp$baseline, 
                    trainingBL = trainResp$baseline,
                   controlStress = controlResp$stressor,
                   trainingStress = trainResp$stressor)


resp_plot <- melt(resp_plot)
resp_plot$time <- c(rep("baseline", 48), rep("stressor", 48))
resp_plot$group <- c(rep("control", 24), rep("training", 24), rep("control", 24), rep("training", 24))


resp_plot %>% 
  group_by(time, group) %>% 
  summarise(resp_mean = mean(value)) -> melt_resp

resp_plot %>% 
  group_by(group, time) %>% 
  summarise(sd = sd(value)) -> melt_resp_sd

melt_resp$sd = melt_resp_sd$sd
melt_resp$sd = melt_resp$sd / sqrt(24)

rm(melt_resp_sd)


resp_lines <- ggplot(data = melt_resp) +
  aes(x = time, y = resp_mean, color = group) +
  geom_line(aes(group = group), linewidth = 1.2) +
  scale_color_manual(values=cbPalette)+
  labs(title = "",
       x = "Condition",
       y = "Respiration Rate ",
       fill = "group")+
  geom_point() +
  ggtitle("Respiration Change from Baseline to \n Post-Training Stressor")+
  theme_minimal()+
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(plot.title = element_text(margin=margin(0,0,13,0))) +
  theme(plot.title=element_text(face="bold")) +
  theme(plot.title=element_text(size=12)) +
  theme(axis.text = element_text(colour = "#00000095")) +
  theme(axis.title.x = element_text(size = 13)) +
  theme(axis.title.y = element_text(size = 14)) +
  theme(axis.text=element_text(size=12))+
  theme(axis.title.x = element_text(margin=margin(13,0,0,0))) +
  theme(axis.title.y = element_text(margin=margin(0,13,0,0))) +
  theme(legend.text = element_text(size=18))+
  theme(legend.key.size = unit(2, 'cm'))+
  geom_errorbar(aes(ymin = resp_mean-sd, ymax= resp_mean+sd),width=.2,
                position=position_dodge(0.05)) +
  scale_y_continuous(n.breaks = 7)

resp_lines
```
